Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-11 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
add case study III and update section 4.8

%In the online supplemental material, additional case studies are presented. From those examples, it can be seen how 
The computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})).
In the experiments of this paper computation for HSGP model was faster than for the exact GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material).
For small and moderate sized datasets, HSGPs is likely to be slower than exact GPs already for $D=3$ and $D=4$ even with a relatively low number of basis functions ($m \gtrsim 5$ and $m \gtrsim 20$ respectively; see Figure~C.3 in the online material). The HSGP method can be computationally faster than the exact GP for larger datasets due the cubic scaling of the exact GP. In all of the investigated cases, choosing the optimal booundary factor $c$ reduces the required number of basis functions $m$ noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore  reduces computational time drastically in particular in multidimensional cases. When HSGP or exact GP is faster depend also on the implementation details that can have big effect on the constants, and thus any more detailed recommendations can only be made in the context of the specific software implementation.

@@

The HSGP model is fitted following the recommended iterative steps as in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on $f_1$, $f_2$ and $f_3$, where for each one these functions the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated. For functions $f_2$ and $f_3$ there are not boundary factor $c$ as they use periodic covariance functions, and $m$ and minimum $\ell$ are updated by eq.~\eqref{eq_J_l_periodic}. A initial value for the minimum lengthscale $\ell_1$ of $f_1$ is guessed to correspond to around 3 years (i.e., $\ell_1=3$ years $= 3\cdot 365=1095$ days) of the input dimension. Initial values for the minimum lengthscales $\ell_2$ and $\ell_3$ of $f_2$ and $f_3$, respectively, are guessed to correspond to half of the period (i.e., $l_2=l_3=0.5$).

\begin{table}
\centering
\setlength{\tabcolsep}{4pt}
\begin{tabular}{ccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
iter. & f & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & ELPD \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.52 & 1.20 & 7  & 0.53 & FALSE &      &  	   \\
 1 & $f_2$ & 0.50 & -    & 8  & 0.34 & FALSE & 0.32 & 0.05 \\
   & $f_3$ & 0.50 & -    & 8  & 1.62 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.50 & 1.20 & 8  & 0.54 & TRUE  &      &  	   \\
 2 & $f_2$ & 0.34 & -    & 11 & 0.29 & FALSE & 0.32 & 0.06 \\
   & $f_3$ & 0.29 & -    & 13 & 1.53 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 13 & 0.33 & TRUE  &      &  	   \\
 3 & $f_2$ & 0.29 & -    & 13 & 0.24 & FALSE & 0.30 & 0.12 \\
   & $f_3$ & 0.29 & -    & 13 & 1.64 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 18 & 0.23 & TRUE  &      &  	   \\
 4 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.30 & 0.15 \\
   & $f_3$ & 0.29 & -    & 13 & 1.68 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.16 & 1.20 & 23 & 0.20 & TRUE  &      &  	   \\
 5 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.29 & 0.18 \\
   & $f_3$ & 0.29 & -    & 13 & 1.63 & TRUE  &      &  	   \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{Iterative steps in HSGP model fitting of \textit{Birthday data} case study. The diagnosis procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three underlying functions $f_1$, $f_2$ and $f_3$. Notice that $f_2$ and $f_3$ do not have boundary factor $c$ as they use periodic covariance functions. RMSE and ELPD are evaluated for the model underlying function $\mu=f_1 + f_2 + f_3 + f_4$ and sample data. The full diagnosis process is applied until two \textit{trues} are achieved for each function.}
  \label{tab_caseI}
\end{table}

@@

\subsection{Case study of simulated data for a bivariate function}

This example consists of a synthetic dataset with $n=120$ ($i=1,\dots,n$) single draws from a GP prior as a function of input values $\bm{x}_i \in \{[-1,1],[-1,1]\} \subset {\rm I\!R}^2$ in a bivariate input space ($D=2$). A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, for the first input dimension, and $\ell_2=0.35$, for the second input dimension, is used for the GP prior. Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy set of observations $\bm{y} \in {\rm I\!R}^{120}$.

In the HSGP model with $2$ input dimensions, the latent function $f(\bm{x})$ is approximated as in equation (13)% (\ref{eq_approxf_multi})
, with the $2$D squared exponential spectral density $S$ as in eq. \eqref{eq_specdens_inf}, and the $2$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations \eqref{eq_eigenfunction_multi} and \eqref{eq_eigenvalue_multi}, respectively.

The HSGP model is fitted following the recommended iterative steps as in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on both dimensions, where for each one the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated by equations in Section~\ref{sec_num_equations}. The initial values $0.5$ and $1$ are set for the minimum lengthscales $\ell_1$ and $\ell_2$, for first and second dimensions respectively.

In order to do model comparison, in addition to the exact GP and HSGP models, a $2$D splines-based model is also fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig15_Posteriors_exII} shows the data-generating GP function, from where the dataset was drawn, and the mean posterior predictive functions of the three models, exact GP, HSGP, and splines, fitted over the dataset. Sample observations are also plotted as circles. For the HSGP, $m_1=40$ and $m_2=40$ basis functions for each dimension, respectively, were used, which lead to a total of 1600 multivariate basis functions. A boundary factor for each dimension, $c_1=1.5$ and $c_2=1.5$, were used. For the spline-based model, 40 knots in each dimension were used.

Figure \ref{ch5_fig16_errors_exII} shows the difference functions between the data-generating function and the GP, HSGP and spline-based models, respectively.

\begin{figure*}
\centering
\includegraphics[scale=0.36, trim = 1mm 18mm 39.2mm 20mm, clip]{ch5_fig15_gpfun_exII.pdf}
\includegraphics[scale=0.36, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig15_bffun_exII.pdf}
\includegraphics[scale=0.36, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig15_spfun_exII.pdf}
\includegraphics[scale=0.30, trim = 150mm 5mm 0mm 20mm, clip]{ch5_fig15_colorbar_vertical_exII.pdf}
\caption{Data-generating function (a) and posterior predictive mean functions of the GP (b), HSGP (c) and spline-based (d) models. Sample points are plotted as circles}
  \label{ch5_fig15_Posteriors_exII}
\end{figure*}

\begin{figure*}
\centering
\subfigure{\includegraphics[scale=0.36, trim = 1mm 18mm 39.2mm 20mm, clip]{ch5_fig16_errorGP_exII.pdf}}
\subfigure{\includegraphics[scale=0.36, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig16_errorHSGP_exII.pdf}}
\subfigure{\includegraphics[scale=0.36, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig16_errorSP_exII.pdf}}
\subfigure{\includegraphics[scale=0.30, trim = 150mm 5mm 0mm 20mm, clip]{ch5_fig16_error_colorbar_exII.pdf}}
\caption{Error between the data-generating function and posterior mean functions of the GP (a), HSGP (b) and spline-based (c) models. Sample points are plotted as circles.}
  \label{ch5_fig16_errors_exII}
\end{figure*}

In order to assess performance of the models as a function of the number of basis functions, boundary factor and knots, different models with different number of basis functions, for the HSGP, and different number of knots, for the spline-based model, have been fitted. In each of these models, the same boundary factor, number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP, and knots, for the spline-based model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-left, it can be seen an accurate approximation of the HSGP to the exact GP. However, the performance of the spline-based model is significantly worse.

Figure \ref{ch5_fig17_RMSE_exII}-right shows the computation times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computation time. Even though in a bivariate input space the computational demands increase significantly with the number of dimensions or knots, either the HSGP or spline-based models work significantly better than exact GP, even for highly wiggly functions that require a high number of basis functions or knots for an accurate approximation. It is interesting to be noticed that just the computation of the input data for the spline model in a 2D input space and considering more than 50 knots per dimension is computationally very expensive.

>>>>

The computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). The HSGP method can be computationally faster than the exact GP for larger datasets due the cubic scaling of the exact GP.

The HSGP model is definitely useful in $1$D and $2$D, even for highly wiggly functions, unless in $2$D the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already reasonably fast itself.

In the experiments of this paper for multivariate cases, mostly reported in Section~\ref{sec_computation_nD}, computation for HSGP model was faster than for the exact GP for most of the non-linear $2D$ functions or moderated-large sized ($n>1000$) $3$D datasets (even highly wiggly functions with large $m^* \approx 10000$); see Figure~\ref{fig_time_nD_log}).

For small sized ($n<1000$) datasets, HSGPs is likely to be slower than exact GPs already for $D=3$ and $D=4$, with a relatively large number of multivariate basis functions ($m^* > 3000$) in $3$D, and with a relatively low number of multivariate basis functions ($m^* \gtrsim 3000$) in $4$D (see Figure~\ref{fig_time_nD_log}).

%In all of the investigated cases, choosing the optimal booundary factor $c$ reduces the required number of basis functions $m$ noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore  reduces computational time drastically in particular in multidimensional cases.

As it has been shown in case study III (Section~\ref{sec_caseIII}), the iterative diagnostic tool proposed in this study can be very useful for the multivariate cases as it allows to reduce $m^*$ to the minimum, reducing computational time drastically, and still getting an accurate approximation. Otherwise, user might be struggling in these multivariate cases.

When HSGP or exact GP is faster depend also on the implementation details that can have big effect on the constants, and thus any more detailed recommendations can only be made in the context of the specific software implementation.

@@

The HSGP model is fitted following the recommended iterative steps as in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on $f_1$, $f_2$ and $f_3$, where for each one these functions the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated. For functions $f_2$ and $f_3$ there are not boundary factor $c$ as they use periodic covariance functions, and $m$ and minimum $\ell$ are updated by eq.~\eqref{eq_J_l_periodic}. A initial value for the minimum lengthscale $\ell_1$ of $f_1$ is guessed to correspond to around 3 years (i.e., $\ell_1=3$ years $= 3\cdot 365=1095$ days) of the input dimension. Initial values for the minimum lengthscales $\ell_2$ and $\ell_3$ of $f_2$ and $f_3$, respectively, are guessed to correspond to half of the period (i.e., $l_2=l_3=0.5$). After the diagnostic generated the first \textit{true}, $m$ is updated by increasing the $m$ of the previous iteration by 5 additional basis functions and $c$ and the minimum $\ell$ updated accordingly as explained in Section~\ref{sec_user_guide}. The full diagnosis process is applied until two \textit{trues} are achieved for each function. Table~\ref{tab_caseII} contains the values for the parameters $\ell$, $c$, $m$, the estimated $\hat{\ell}$, the diagnostic $\hat{\ell} + 0.01 \geq \ell$ for each function, and the RMSE compute with the data as the reference for every iterative steps of the fitting process.

\begin{table}
\centering
\setlength{\tabcolsep}{4pt}
\begin{tabular}{ccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
iter. & f & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & ELPD \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.52 & 1.20 & 7  & 0.53 & FALSE &      &  	   \\
 1 & $f_2$ & 0.50 & -    & 8  & 0.34 & FALSE & 0.32 & 0.05 \\
   & $f_3$ & 0.50 & -    & 8  & 1.62 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.50 & 1.20 & 8  & 0.54 & TRUE  &      &  	   \\
 2 & $f_2$ & 0.34 & -    & 11 & 0.29 & FALSE & 0.32 & 0.06 \\
   & $f_3$ & 0.29 & -    & 13 & 1.53 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 13 & 0.33 & TRUE  &      &  	   \\
 3 & $f_2$ & 0.29 & -    & 13 & 0.24 & FALSE & 0.30 & 0.12 \\
   & $f_3$ & 0.29 & -    & 13 & 1.64 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 18 & 0.23 & TRUE  &      &  	   \\
 4 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.30 & 0.15 \\
   & $f_3$ & 0.29 & -    & 13 & 1.68 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.16 & 1.20 & 23 & 0.20 & TRUE  &      &  	   \\
 5 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.29 & 0.18 \\
   & $f_3$ & 0.29 & -    & 13 & 1.63 & TRUE  &      &  	   \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{Iterative steps in HSGP model fitting of \textit{Birthday data} case study. The diagnosis procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three underlying functions $f_1$, $f_2$ and $f_3$. Notice that $f_2$ and $f_3$ do not have boundary factor $c$ as they use periodic covariance functions. RMSE and ELPD are evaluated for the model underlying function $\mu=f_1 + f_2 + f_3 + f_4$ and sample data. The full diagnosis process is applied until two \textit{trues} are achieved for each function.}
  \label{tab_caseII}
\end{table}

@@

\subsection{Case study of simulated data for $2$D and $3$D functions}\label{sec_caseIII}

In this example we apply the diagnostic tool to fit and diagnose two different data sets, one data set simulated from a bivariate (D=$2$) function and another data set simulated from a $3$-dimensinal (D=$3$) function. Furthermore, we present results about the computational time required to fit the HSGP model in $2$D, $3$D and $4$D input spaces, with different sized data sets and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation.

The $2$D and $3$D synthetic datasets consist of $n=200$ and $n=1000$ ($i=1,\dots,n$) single draws from GP priors with input values $\bm{x}_i \in \{[-1,1],\dots,[-1,1]\} \subset {\rm I\!R}^D$, in a D-dimensional input space. A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and length-scales $\ell_1=0.10$, for the first input dimension, $\ell_2=0.3$, for the second input dimension, and $\ell_3=0.4$, for the third input dimension, is used for the GP priors. Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy sets of observations.

In the HSGP models with $2$ and $3$ input dimensions, the underlying functions $f(\bm{x})$ to the data are approximated as in eq. (13)% (\ref{eq_approxf_multi})
, with the $D$-dimensional squared exponential spectral densities $S$ as in eq. \eqref{eq_specdens_inf}, and the $D$-vectors of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations \eqref{eq_eigenfunction_multi} and \eqref{eq_eigenvalue_multi}, respectively.

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,3)$ and $\mathrm{InverseGamma}(2,0.5)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variance $\alpha$, and length-scales $\ell$, respectively. We use the same prior distributions to fit the exact GP model as the HSGP models on the synthetic data sets.

The HSGP models are fitted following the recommended iterative steps as explained in Section~\ref{sec_user_guide}, where in each iteration the diagnosis is applied on every dimensions and, for each one of the dimensions, the parameters $c$, $m$, minimum $\ell$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ are updated by equations in Section~\ref{sec_num_equations}. The values $0.5$, $1$, and $1$ are set as initial values for the minimum lengthscales $\ell_1$, $\ell_2$ and $\ell_3$, for the first, second and third dimensions, respectively. In order to be as more efficient as possible in these multivariate cases, after the diagnostic generated the first \textit{true} for a certain dimension, its corresponding $m$ is updated by increasing the $m$ of the previous iteration by only 2 additional basis functions and, after the second \textit{true}, $m$ is no longer increased. The full diagnosis process is applied until two \textit{trues} are achieved for each dimension.

Table~\ref{tab_caseIII_2D} and \ref{tab_caseIII_3D} contain the iterative steps to fit and diagnose the $2$D and $3$D data sets, respectively. 
%These tables contain the values for the parameters $\ell$, $c$, $m$, the estimated $\hat{\ell}$, the diagnostic $\hat{\ell} + 0.01 \geq \ell$ for each dimension, and the RMSE compute with both the data and the GP as the reference for every iterative steps of the fitting process. 
The minimum requirements to fit the models were easily achieved by performing 4 iterations: from a $1^{st}$ iteration that uses few multivariate basis functions $m^*$ ($m^*=m_1(=6) \times m_2(=6)=36$ and $m^*=m_1(=6) \times m_2(=6) \times m_3(=6)=216$ for the $2$D and $3$D data sets, respectively) to the $4^{th}$ iteration that uses the minimum multivariate basis functions ($m^*=m_1(=22) \times m_2(=11)=242$ and $m^*=m_1(=21) \times m_2(=10) \times m_3(=12)=2520$ for the $2$D and $3$D data sets, respectively) required to accurately fit these notably wiggly data sets. Without using our recommended guidelines and diagnostics, it could have taken user several trials to get the optimal solution and to expend a significant amount of time, as using a larger number of basis functions in multivariate cases increases time of computation drastically.

\begin{figure*}[h]
\centering
\subfigure{\includegraphics[scale=0.315, trim = 1mm 18mm 39.2mm 20mm, clip]{ch5_fig16_errorGP_exII.pdf}}
\subfigure{\includegraphics[scale=0.315, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig16_errorHSGP_exII.pdf}}
\subfigure{\includegraphics[scale=0.315, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig16_errorSP_exII.pdf}}
\subfigure{\includegraphics[scale=0.30, trim = 151mm 14mm 2mm 20mm, clip]{ch5_fig16_error_colorbar_exII.pdf}}
\hspace{1pt}\subfigure{\includegraphics[scale=0.27, trim =0mm 5mm 10mm 0mm, clip]{ch5_fig17_RMSE_exII.pdf}}
\hspace{-1.4cm} \subfigure{\raisebox{1.65cm}{\includegraphics[scale=0.30, trim = 28mm 70mm 112mm 20mm, clip]{ch5_fig17_legend_exII.pdf}}}
\caption{Error between the $2$D data-generating function and posterior mean functions of the GP (a), HSGP (b) and spline-based (c) models. Sample points are plotted as circles. The right side plot shows the root mean square error (RMSE) of the different methods, and plotted as a function of the boundary factor $c$, number of univariate basis functions $m$ and knots (the same number of basis functions $m$ and knots per dimension are used, resulting in $m^D$ total number of multivariate basis functions).}
  \label{ch5_fig16_errors_exII}
\end{figure*}

\begin{table}
\centering
\setlength{\tabcolsep}{3.6pt}
\begin{tabular}{cccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
 iter. & Dim. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 1 } & $1$ & 0.50 & 1.6 & 6 & 0.19 & FALSE & \multirow{2}{*}{ 0.75 } & \multirow{2}{*}{ 0.72 }\\ 
 & $2$ & 1.00 & 3.2 & 6 & 0.25 & FALSE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 2 } & $1$ & 0.19 & 1.2 & 12 & 0.11 & FALSE  & \multirow{2}{*}{ 0.18 } & \multirow{2}{*}{ 0.12 }\\
 & $2$ & 0.25 & 1.2 & 9 & 0.25 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 3 } & $1$ & 0.11 & 1.2 & 20 & 0.11 & TRUE  & \multirow{2}{*}{ 0.11 } & \multirow{2}{*}{ 0.03 }\\
 & $2$ & 0.19 & 1.2 & 11 & 0.27 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{2}{*}{ 4 } & $1$ & 0.10 & 1.2 & 22 & 0.11 & TRUE  & \multirow{2}{*}{ 0.10 } & \multirow{2}{*}{ 0.02 }\\
 & $2$ & 0.19 & 1.2 & 11 & 0.27 & TRUE  &  & \\[1mm] 
\arrayrulecolor{lightgray}\hline
 \multicolumn{8}{l}{ $\ell_{1_{GP}}=0.10$, \, $\ell_{2_{GP}}=0.29$ }
\end{tabular}
\caption{Iterative steps in HSGP model fitting on the $2$D data. The diagnosis procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the two dimensions $dim_1$ and $dim_2$. The full diagnosis process is applied until two consecutive \textit{trues} in the diagnostic are achieved for each dimension. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab_caseIII_2D}
\end{table}

\begin{table}
\centering
\setlength{\tabcolsep}{3.6pt}
\begin{tabular}{cccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
 iter. & Dim. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & RMSE* \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 1 } & $1$ & 0.50 & 1.6 & 6 & 0.14 & FALSE & \multirow{3}{*}{ 0.63 } & \multirow{3}{*}{ 0.58 }\\ 
 & $2$ & 0.50 & 1.6 & 6 & 0.27 & FALSE &  &  \\
 & $3$ & 1.00 & 3.2 & 6 & 0.23 & FALSE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 2 } & $1$ & 0.13 & 1.2 & 16 & 0.11 & FALSE  & \multirow{3}{*}{ 0.13 } & \multirow{3}{*}{ 0.07 }\\
 & $2$ & 0.27 & 1.2 & 8 & 0.29 & TRUE &  &  \\
 & $3$ & 0.23 & 1.2 & 19 & 0.40 & TRUE &  &  \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 3 } & $1$ & 0.11 & 1.2 & 19 & 0.12 & TRUE  & \multirow{3}{*}{ 0.11 } & \multirow{3}{*}{ 0.04 }\\
 & $2$ & 0.21 & 1.2 & 10 & 0.29 & TRUE &  &  \\
 & $3$ & 0.19 & 1.3 & 12 & 0.43 & TRUE &  &  \\
 \arrayrulecolor{lightgray}\hline \\[-2mm]
 \multirow{3}{*}{ 4 } & $1$ & 0.10 & 1.2 & 21 & 0.12 & TRUE  & \multirow{3}{*}{ 0.11 } & \multirow{3}{*}{ 0.03 }\\
 & $2$ & 0.21 & 1.2 & 10 & 0.29 & TRUE &  &  \\
 & $3$ & 0.19 & 1.3 & 12 & 0.43 & TRUE &  &  \\[1mm] 
\arrayrulecolor{lightgray}\hline
 \multicolumn{9}{l}{ $\ell_{1_{GP}}=0.11$, \,$\ell_{2_{GP}}=0.32$, \, $\ell_{3_{GP}}=0.43$ }
\end{tabular}
\caption{Iterative steps in HSGP model fitting on the $3$D data. The diagnosis procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three dimensions $dim_1$, $dim_2$ and $dim_3$. The full diagnosis process is applied until two consecutive \textit{trues} in the diagnostic are achieved for each dimension. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab_caseIII_3D}
\end{table}

%\begin{figure*}
%\centering
%\includegraphics[scale=0.36, trim = 1mm 18mm 39.2mm 20mm, clip]{ch5_fig15_gpfun_exII.pdf}
%\includegraphics[scale=0.36, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig15_bffun_exII.pdf}
%\includegraphics[scale=0.36, trim = 19mm 18mm 39.2mm 20mm, clip]{ch5_fig15_spfun_exII.pdf}
%\includegraphics[scale=0.30, trim = 150mm 5mm 0mm 20mm, clip]{ch5_fig15_colorbar_vertical_exII.pdf}
%\caption{Data-generating function (a) and posterior predictive mean functions of the GP (b), HSGP (c) and spline-based (d) models. Sample points are plotted as circles}
%  \label{ch5_fig15_Posteriors_exII}
%\end{figure*}

%\begin{figure}
%\centering
%\subfigure{\includegraphics[scale=0.2, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig17_RMSE_exII.pdf}}
%\subfigure{\includegraphics[scale=0.2, trim = 0mm 0mm 10mm 0mm, clip]{ch5_fig17_time_exII.pdf}}
%\hspace{-1.8cm} \subfigure{\raisebox{0.5cm}{\includegraphics[scale=0.30, trim = 28mm 70mm 112mm 20mm, clip]{ch5_fig17_legend_exII.pdf}}}
%\caption{Root mean square error (RMSE) (left) and computational time (right) in seconds per iteration (iteration of the HMC sampling method) of the different methods, and plotted as a function of the boundary factor $c$, number of basis functions $m$ and knots.}
%  \label{ch5_fig17_RMSE_exII}
%\end{figure}

%Figure \ref{ch5_fig15_Posteriors_exII} shows the data-generating GP function, from where the dataset was drawn, and the mean posterior predictive functions of the three models, exact GP, HSGP, and splines, fitted over the dataset. Sample observations are also plotted as circles. For the HSGP, $m_1=40$ and $m_2=40$ basis functions for each dimension, respectively, were used, which lead to a total of 1600 multivariate basis functions. A boundary factor for each dimension, $c_1=1.5$ and $c_2=1.5$, were used. For the spline-based model, 40 knots in each dimension were used.

Figure \ref{ch5_fig16_errors_exII} shows the difference functions between the $2$D data-generating function, from where the dataset was drawn, and the GP, HSGP and spline-based model solutions, respectively. The $2$D splines-based model was fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline-based model using the R-package \textit{brms} \citep{burkner2017brms}. For the spline-based model, $40$ knots in each dimension were used. Figure \ref{ch5_fig16_errors_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$ and number of univariate basis functions $m$, for the HSGP, and knots, for the $2$D spline-based model. It can be seen an accurate approximation of the HSGP and exact GP to the $2$D data-generative function. However, the performance of the spline-based model is significantly worse.

\subsubsection{Computation requirements in $2$D, $3$D and $4$D input spaces}\label{sec_computation_nD}

Experiments have been conducted to empirically recognize time of computation of the HSGP model in multivariate cases and different number of sample data points. Dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual} was used for the experiments. Implementation and sampling has been done only for the sample data.

Figure~\ref{fig_time_nD_log} shows the computation time (in seconds per iteration) in $2$D, $3$D and $4$D input spaces and different sized data sets, $n=300, n=500, n=1000$ and $n=3000$, and as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$ used in the approximation. Both the time of computation and $m^*$ are plotted in the logarithm scale.

If we look back at Figure~\ref{fig6_relationships} or equations~\eqref{eq_m_l_QE}, \eqref{eq_m_l_mat32} and \eqref{eq_m_l_mat52} we can find out that, for example for a squared exponential kernel, any univariate function (or single dimensions of a multivariate function) with true lengthscales bigger than $0.3$ can be accurately fitted, after applying the iterative diagnostic steps, by only using approximately $10$ or $12$ basis functions as much, with the assumption that we have chosen the optimal value of the boundary factor $c$. And, for true lengthscales between $0.1$ and $0.3$, by only using between $10$ and $22$ basis functions as much, with the assumption that we have chosen the optimal value of the boundary factor $c$.

So, for a very wiggly $2$D function (let's say with $\ell_1=0.1$ and $\ell_2=0.1$), the approximate number of multivariate basis functions needed for an accurate approximation will be $m*=22\times 22= 484$, which results in an approximation significantly faster than the regular GP even in very small sized data sets (i.e., $n<300$) (see Figure~\ref{fig_time_nD_log}). Similarly, for a very wiggly $3$D function (let's say with $\ell_1=0.1$, $\ell_2=0.1$ and $\ell_3=0.1$), we can expect an approximate minimum number of multivariate basis functions for an accurate approximation of $m*=22\times 22\times 22= 10648$, which, unless it is a small sized data set (i.e., $n<1000$), the method is significantly faster than the regular GP. Note the a $3$D function where each dimension has a lengthscale around $0.1$ is quite uncommon, so for most of the $3$D functions, unless it is a very small sized data set, the approximation will be significantly faster than the excat GP, if our iterative diagnostic rules are used as they allow to find out easily the solution with the minimum requirements. For $4$D data sets, the method can still be more efficient than the regular GP for moderate-large data sets (i.e., $n>1000$).

Finally, for $D>5$ the method starts to be impractical even for smooth univariate functions and small $n$. However, in these cases the method may still be used in an additive modeling scheme with low-dimensional additive components.

%Even though in a bivariate input space the computational demands increase significantly with the number of dimensions or knots, either the HSGP or spline-based models work significantly better than exact GP, even for highly wiggly functions that require a high number of basis functions or knots for an accurate approximation.

%\begin{figure*}
%\centering
%\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig_time_nD.png}
%\caption{Computational time in seconds per iteration (iteration of the HMC sampling method) of the different methods, and plotted as a function of the boundary factor $c$, number of multivariate basis functions $m^D$ and for different number of data points.}
%  \label{fig_time_nD}
%\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig_time_nD_log.png}
\caption{Computational time in seconds per iteration (iteration of the HMC sampling method) for different datasets with different dimensionalties and data points, and plotted as a function of the number of multivariate basis functions $m^*=m_1\times \cdots \times m_D$.}
  \label{fig_time_nD_log}
\end{figure*}
